### 2018-01-11 paper/response_to_reviewers.tex

\setlength\parindent{0pt}

\begin{document}

\textbf{RE: MEE-17-08-729 }\\
\bigskip
Dear Dr Vamosi,\\
\bigskip

I am very grateful to the four referees for their helpful and constructive comments, which I believe have helped me to significantly improve the \texttt{dispRity} package and this associated package description paper.
I have taken all of their comments on board (though some will be implemented in later versions), and respond to their points in details below.
For improving clarity in this document, reviewers' comments are displayed in blue, my responses in normal text and the changes to the manuscript in italic.
Additionally, I've uploaded two versions of the revised manuscript: both have the exact same content but one has all the changes and additions to the main text highlighted in yellow to help with reviewing.
All the changes implemented in the package and the vignette are available on the release branch under version 0.5.1.

@@

\begin{enumerate}

\item{\textcolor{blue}{While I understand that the author is writing a package that has many diverse capabilities and is presented to be useful to a wide range of disciplines, it is too vague in too many places to be helpful in its current state.
For example, the line in the abstract “It is then possible to use a summary metrics for measuring some properties of this space for some group of species and compare them.”.
“Some properties” is too vague without any qualifiers to be useful. But I believe that with revision, taking into account the issues of clarity, this will be an important addition to the available analytical tools for Biologists.}}

I have changed the terms used in the abstract and proposed more specific examples for clarity:

\textit{It is then possible to measure diverse aspects of the distribution of some observation (e.g. species) in this space. For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.} lines 6-10.

@@

The changes have also been made throughout the manuscript as generally suggested by all four reviewers.

\item{\textcolor{blue}{My main concern is that what disparity actually means is left to the reader’s imagination – it is not defined or discussed anywhere.
Disparity simply means lack of similarity or equality.
While the author is probably considering the dispersion of points in a multidimensional space as a measure of dissimilarity, it is important to realise the ambiguity of the word and more explanation is needed early in the introduction (and foremost in the abstract).
Specifically, the lines 58-61 need to be revised, as they simply don’t convey to the reader what is being presented here as disparity.
Say what “specific aspect of it [multidimensional space]” you are referring to.
Also, the different definitions of disparity alluded to on lines 61-63 should be summarised.
Furthermore, the manuscript is lacking a table summarising the metrics of disparity that this package can work with (which I suspect it was supposed to have).}}
\label{define_disparity}

Apologies for the omission of the table summarising the metrics which is now properly included in the manuscript (see reviewer 1, minor comment \ref{table_missing}).
As reviewer 4 (major comment \ref{nebulous}) also underlines: ``disparity is a broad, nebulous concept''.
This observation was the genesis for this package and I am currently working on a collaborative review to define the borders (or the absence thereof) of this concept.
I feel that explaining this would be out of the scope of this software description paper.
However, I have tried to give specific examples each time I refer to disparity as a way to measure these ``specific aspect of the space'' and I have also added a clearer definition of what I mean by disparity throughout this manuscript.

@@

\item{\textcolor{blue}{Another concern I have is about how this R package sits within the current analytical toolkit available.
How does this new package relate to existing packages available for R?
Claddis and geomorph are mentioned and discussed because they deal with bespoke data formats that need to be dealt with prior to analysis.
But are there others?
How does the package relate to the ‘vegan’ R package for multivariate analysis of ecological communities?
Within there are metrics for community dissimilarity, metrics from ordinations etc.
Vegan is a popular package for analysis of multivariate datasets and seems to be relevant here with similar functionalities but not discussed.
Similarly, the R package ‘diverse’ has a function called disparity, and the ‘geiger’ function disparity implements the disparity-through-time. }}
\label{add_geiger_vegan}

I have added two wrapper functions to use the popular functions \texttt{vegan::adonis} and \texttt{geiger::dtt} to the package (\texttt{dispRity.adonis} and \texttt{dispRity.dtt}).
Also, I have generalised the distances calculations using the \texttt{vegan::vegdist} function that allows for more distances than \texttt{stats::dist} to be calculated.

I've updated the manual and the example to describe these new functions:

\textit{Similarly to the example above, it is also possible to statistically test this hypothesis using, for example, permutational multivariate ANOVA (PERMANOVA; Anderson, 2001b) through the adonis.disprity function that is a wrapper of the vegan::adonis function (Oksanen et al., 2007) for dispRity objects [TABLE 4]. To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset. Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test. Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.} lines 359-369.

@@

\item{\textcolor{blue}{Line 4 – changes ``in'' to ``to''. Line 6 – changes ``axis'' to ``axes''}}
\label{abstract_typo}

I have fixed these typos.

\item{\textcolor{blue}{Line 84 – Explain what the Procrustes variance is. And when is it appropriate and when is not appropriate? Be more specific.}}
\label{tone_down}

In \texttt{geomorph}, the procrustes variance is defined as the sum of the diagonal covariance matrix between each elements divided by the number of elements (or effectively the sum of the squared ordination values divided by the number of elements - see example in \texttt{R} for \texttt{?dispRity::geomorph.ordination}).
Using this disparity metric is appropriate for testing volume based hypothesis (e.g. Does group A occupy more morphospace than group B?) but does not allow to test position based hypothesis (e.g. Does group A occupy a different part of the morphospace than group B).
For example, one might find out that group A and B have the same volume but that doesn't tell us if they occupy the same position in the morphospace.

@@

\item{\textcolor{blue}{Lines 98-100 – Suggest changing this to be specific: “calculates a disparity metric from the space”, and “analyse the resulting disparity object through hypothesis testing and visualisation”}}

I have changed the sentence following the reviewer's suggestion.

\item{\textcolor{blue}{Line 116 – “Table ??” = should have been caught during proofreading. Furthermore, there is no table included in the manuscript that details the implemented metrics, and this is an important part of the paper sadly omitted in this submission.}}
\label{table_missing}

I sincerely apologise for the omission of the Table! It is now included in the new version of the manuscript with the currently implemented metrics description.
I would also like to underline that although not all usual disparity metrics are represented in the table and in the package (by far!) the idea of the \texttt{dispRity} package and specifically the \texttt{metric} argument in the \texttt{dispRity} function is to allow users to come up with their \textit{own} metric.

@@

I have however modified the available models to better describe this limitation and I have added two new models that would approximate ancestral reconstruction along edges.
I have modified the name of the two previous ``punctuated'' and ``gradual'' models to ``random'' and ``proximity'' to reflect more accurately how they select either the ACCTRAN or DELTRAN information.
I've added two more accurate models (now named ``equal.split'' and ``gradual.split'') to reflect the changes along the branches.
These two algorithm now allows to use both the ACCTRAN and DELTRAN information with a linked probability to use either of them per bootstrap draw.
For example, the ``equal.split'' model will have a 0.5 probability of using either node information for each bootstrap draw; the ``gradual.split'' model transforms the branch length were the slice occurs into a probability of drawing either of the nodes information (e.g. if the slices occurs 3/4 along the branch from A to B, the model will draw the information from A and B with a probability of respectively 0.25 and 0.75).
For example, the median disparity over 1000 bootstrap pseudo-replication along the edge (AB) will integrate 0.75 information from A and 0.25 from B thus approximating the ``intermediate'' between A and B (when averaged ).

I've added these models descriptions in the manual and in the manuscript:

\begin{itemize}
    \item \textit{``random'' where the data chosen along the branch is randomly chosen between the descendant or the ancestor}
    \item \textit{``proximity'' where the data chosen along the branch is either the descendant or the ancestor depending on branch length}
    \item \textit{``equal.split'' where the data chosen along the branch is both the descendant and the ancestor with an even probability}
    \item \textit{``gradual.split'' where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length}
\end{itemize}

\textit{Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants (see below).
Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} lines 169-183.

@@

I also agree with this comment: this method is a trade-off between the number of slices.
(1) too few slices decrease resolution and veer towards the time binning (which could increase "significance" by comparing too different time subsamples);
(2) too many slices will increase the resolution be will create much more pseudo replication and thus increase type I error.

I've added a warning in the manual and the manuscript:

\textit{Finally, there is a trade-off between precision and accuracy when using the time-slicing method: a higher number of slices increases the precision of the disparity analysis but also increases the type I error (thus decreasing accuracy).} line 180-183.

A more detailed description and test of these models might be out of the scope of this package description paper but is currently in development for a special issue paper in \textit{Palaeontology} (in review) as part as a special symposium for the Palaeontological Association annual meeting 2017 in collaboration with Natalie Cooper.
The full (un-reviewed an un-compiled) manuscript is available on Natalie's Github page (\url{https://github.com/nhcooper123/time-slice/blob/master/manuscript/time-slice.tex}).

@@

\item{\textcolor{blue}{Finally, additions the author could consider adding to the package. These are mostly minor, but more numerous. Some of these are simple caveats for limitations to the package. For example, disparity metrics can be generated from, for example, a discrete-character taxon matrix directly, or the distance matrix given to the ordination (e.g., the weighted mean pairwise distance metric used in Close et al. 2015, cited). Because the package only deals with matrices generated by the ordination itself these metrics are beyond its scope, although are of potential value to disparity analysis more broadly. Specifically, the Close et al. (2015) metric can be applied to an incomplete matrix, but principal coordinates/classic multidimensional scaling requires a complete distance matrix and hence taxa would have to be pruned to generate an ordination, but can be retained if using a pre-ordination metric. In a similar vein, although the package allows bootstrapping of the space it could also be useful for the raw data itself to be bootstrapped, e.g., the set of measurements used to generate a principal components analysis. It might be helpful for the author just to make it clear (with a sentence or two) that these are things a user might want to attempt, but the package cannot (reasonably) do.}}

I have updated the package's structure to be able to deal with any type of matrix as long as rows are the elements (e.g. taxa).
It is now possible to perform the whole pipelined analysis (including bootstrapping) on non-ordinated matrices.

I've reflected these changes in the manual with a subsection in the chapter ``Specific tutorials - 4.8'' (based on a distance matrix) and in manuscript as follows:

\textit{Note that the input matrix is not restricted to an ordinated matrix but can be any kind of matrix as long as its rows represent elements (e.g. the multidimensional space can be a distance matrix; Close et al., 2015).} lines 109-111.

@@

\item{\textcolor{blue}{Picking up on the first point from above, there are also additional post-ordination disparity metrics that have been used in the literature but are not available here. For example, hypercube volume, minimum spanning tree length etc. I would also urge the author to cite the work of Mike Foote, who made some early important contributions to disparity in general, and has written papers that discuss some of these potential metrics (e.g., Foote 1997, Annual Review of Ecology and Systematics, 28, 129–). I believe Wills (2001, cited) also discusses a long list of metrics.}}

I agree, there are many more metrics commonly used for disparity analysis not implemented in this package.
This was a conscious choice: due to the really high number of metrics available (especially when combining different ones), I designed the package to be modular hence allowing users to implement their own metrics (see response to reviewer 1 minor comment \ref{table_missing}).
I have add the hypercube volume description as an example of user defined metric in the manuscript:

\textit{When multiple functions are passed to the} \texttt{metric} \textit{argument, they are sorted by dimension-level and applied in decreasing order to the data.
For example, if the metric is defined as} \texttt{metric = c(prod, ranges)} \textit{(the hypercube volume), the} \texttt{ranges} \textit{function (dimension-level 2) is first applied to data (the range of each dimensions) and the function} \texttt{prod} \textit{is then applied to the results (}\texttt{prod(ranges(data))}\textit{).
One can also directly pass a function description to the} \texttt{metric}\textit{ argument (e.g.} \texttt{metric = function(x) mean(dist(x)$\hat{\mkern6mu}$2)} \textit{ for the average squared pairwise distance).} lines 130-135.

I have however added several additional metrics to the newest version of the package including: (1) the minimal spanning tree length (\texttt{span.tree.length} - as suggested), (2) the ancestral distance which is equivalent to the distance from centroid but were the ``centroid'' are the coordinates of each element's ancestors (\texttt{ancestral.dist}), (3) the pairwise distances (\texttt{pairwise.dist} - based on \texttt{vegan::vegdist}), (4) the radius, equivalent to the distance from centroids but for the dimensions rather than the elements (i.e. the distance between each dimensions coordinates and the mean dimension's coordinates - \texttt{radius}) and (5) the $n$-ball volume which is an extension of the \texttt{ellipse.volume} function for a sphere (the true $n$-ball volume) or an ellipsoid (\texttt{n.ball.volume} - see Table 2 in the new version of the manuscript).

@@

\item{\textcolor{blue}{The package also employs a rarefaction function, but rarefaction can be problematic in the sense that rarefaction curves can sometimes cross such that the sampling level used to make comparisons can effect which of two or more subsets is considered the most disparate. This is why algorithms designed to sample "fairly" rather than equally have been explored, see for example Kotric and Knoll (2015, Paleobiology, 41, 68–). A user may also wish to rarefy by some external data source, such as number of occurrences, or geographic regions. These are obviously major additions, but are something the author might wish to add in future.}}

This is an excellent suggestion for future direction for the package!
I have raised an issue on my GitHub page for implementing the possibility to pass a probability matrix to the \texttt{boot.type} argument in the \texttt{boot.matrix} function where the probabilities can be reflecting occurences, regions, etc.
I will also implement the ``fair'' algorithm mentioned in Kotric and Knoll (2015, Paleobiology, 41, 68–).

### 2018-01-11 paper/dispRity_paper.tex
Response to R1 checked

    \begin{enumerate}
        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
        This causes biologists to either reduce these to a manageable number of variables or, increasingly, to use multivariate toolkits.
        One such toolkit is based on creating a multidimensional space where the variables are the axes.
        It is then possible to measure divers aspects of the distribution of some observation (e.g. species) in this space.
        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.

@@

One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
This requires the definition of a proxy for space occupancy: the disparity metric \citep[or index;][]{Hopkins2017} which can be measured in a multitude of ways.
For example, one could use a metric based on the variance or the range of each axis of space \citep{Wills2001, Ciampaglio2001}, a distance (e.g. Euclidean) measured between observations \citep{foote1993contributions,Foote29111996}, a more direct approximation of the hyper volume \citep{cornwell2006trait,DonohueDim}, or many more \citep[e.g.][]{navarro2003MDA}.

@@

To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.

>>>>

    \begin{enumerate}
        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
        This causes biologists to either reduce these to a manageable number of variables or, increasingly, to use multivariate toolkits.
        One such toolkit is based on creating a multidimensional space where the variables are the axes.
        It is then possible to measure diverse aspects of the distribution of some observation (e.g. species) in this space.
        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not.

@@

One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupy more space than group B?'').
This requires the definition of a proxy for space occupancy: the disparity metric \citep[or index;][]{Hopkins2017} which can be measured in a multitude of ways.
For example, one could use a metric based on the variance or the range of each axis of space \citep{Wills2001, Ciampaglio2001}, a distance (e.g. Euclidean) measured between observations \citep{foote1993contributions,Foote29111996}, a more direct approximation of the hyper volume \citep{cornwell2006trait,DonohueDim}, or many more \citep[e.g.][]{navarro2003MDA}.

@@

To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.

### 2018-01-11 paper/dispRity_paper_highlighted.tex
Response to R1 checked

    \begin{enumerate}
        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
        This causes biologists to either reduce these \hl{to} a manageable number of variables or, increasingly, to use multivariate toolkits.
        One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
        It is then possible to \hl{measure divers aspects of the distribution of some observation (e.g. species) in this space.
        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.

@@

\hl{One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupies more space than group B?'').
This requires the definition of a proxy for space occupancy: the disparity metric}
\citep[or index;][]{Hopkins2017}
\hl{which can be measured in a multitude of ways.}
\hl{For example, one could use a metric based on the variance or the range of each axis of space}
\citep{Wills2001, Ciampaglio2001}
\hl{, a distance (e.g. Euclidean) measured between observations}
\citep{foote1993contributions,Foote29111996}
\hl{, a more direct approximation of the hyper volume }
\citep{cornwell2006trait,DonohueDim}
\hl{, or many more }
\citep[e.g.][]{navarro2003MDA}.

@@

To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
\hl{Note that is this case, the function calls different warnings on the usage of such test and the eventual data not used in the test.
Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.}

>>>>

    \begin{enumerate}
        \item Biological data is multivariate in essence: many traits in organisms co-vary with each other in space and time.
        This causes biologists to either reduce these \hl{to} a manageable number of variables or, increasingly, to use multivariate toolkits.
        One such toolkit is based on creating a multidimensional space where the variables are the \hl{axes}.
        It is then possible to \hl{measure diverse aspects of the distribution of some observation (e.g. species) in this space.
        For example, if studying morphology, one can create a morphospace for two groups of species, measure the volume occupied by each of these groups and then test whether these two volumes are significantly different or not}.

@@

\hl{One can then measure how the observations are distributed within this space to answer related questions (e.g. ``does group A occupy more space than group B?'').
This requires the definition of a proxy for space occupancy: the disparity metric}
\citep[or index;][]{Hopkins2017}
\hl{which can be measured in a multitude of ways.}
\hl{For example, one could use a metric based on the variance or the range of each axis of space}
\citep{Wills2001, Ciampaglio2001}
\hl{, a distance (e.g. Euclidean) measured between observations}
\citep{foote1993contributions,Foote29111996}
\hl{, a more direct approximation of the hyper volume }
\citep{cornwell2006trait,DonohueDim}
\hl{, or many more }
\citep[e.g.][]{navarro2003MDA}.

@@

To answer our specific question above: yes there is an effect of time on morphological disparity (an increase) in this dataset.
\hl{Note that in this case, the function outputs different warnings on the usage of such test and the eventual data not used in the test.
Additionally, the test is not applied to the bootstrapped data and thus might be sensitive to outliers and sampling size.}
